Machine Learning Course Project
========================================================
by Steve Friedman  

Note to graders: A pdf output file has been included in this github repository in case you have trouble viewing the html page online.
# Overview  
In this project we will use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.
The goal of the project is to predict the manner in which they did the exercise.  

````{r}
library(lattice)
library(ggplot2)
library(caret)
```
# Input Data
```{r}
# load the training data
urlfile<-'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
training<-read.csv(urlfile)


#load the testing data
urlfile<-'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'
testing<-read.csv(urlfile)
```
 
# Cleaning Data
A quick visual inspection of the training and testing datasets shows that columns typically have a lot of NA values or hardly any NA values.  We will filter out columns that have 95%  or more of their values set to NA in the training data
```{r}
testing <- testing[, colSums(is.na(training)) < .95*nrow(training) ]
training <- training[, colSums(is.na(training)) < .95*nrow(training) ] 
dim(training); dim(testing)
```
Further inspection shows that there are still a number of columns in the testing data set that have a lot of NA values.
We will now filter out the columns that have 95% or more of their values set to NA in the reduced testing data set
```{r}
training <- training[, colSums(is.na(testing)) < .95*nrow(testing) ] 
testing <- testing[, colSums(is.na(testing)) < .95*nrow(testing) ]
dim(training); dim(testing)
```
We have now filtered the training and testing datasets down to 60 columns that do not have NAs.  The first 7
columns are not directly related to the motion data and will also be removed so the prediction algorithm will work. 
```{r}
training<-training[,-1:-7] # remove columns not related to motion data
testing<-testing[,-1:-7] # remove columns not related to motion data
dim(training);dim(testing)
```
We have now filtered the training and testing datasets down to 53 columns of motion data with no NAs.
# Create Validation Data  
The reduced (no NAs) training dataset will be further partitioned into a second training and a validation dataset to be used for validation of
the model.  Note that the original testing dataset will remain untouched and will not be used until the final predictions 
are done. We will put 70% of this partitioned data into the second training dataset and the other 30% into the validation dataset.
```{r}
set.seed(12345) # make the results reproducible
inTrain <- createDataPartition(training$classe, p=0.70, list=F)
trainData <- training[inTrain, ]
validationData <- training[-inTrain, ]
dim(trainData); dim(validationData)
```
# Create Prediction Model 
A random forest model with 3 fold cross validation will be created.  Random Forest was used because although it is 
computationally intensive it is generally very accurate.  Validation was limited to 3 folds to decrease the computation 
time.
```{r}
controlParams <- trainControl(method="cv",3)
model <- train(classe ~ ., data=trainData, method="rf", trControl=controlParams) 
model
```
# Model Validation  
Predict the outcomes for the validation dataset and then create the confusion matrix to get an idea of the accuracy of 
the predictions.  The confusion matrix as shown directly below looks very accurate.  Most points registered on the 
diagoal (eg A vs A ).  There were a small number (less than 100) of cases where the predicted value differed from the 
actual value in the validation dataset.  The overall accuracy of the model as indicated by the confusion matrix is 98.84%.  Conversly the out of sampe error rate for the validation data set = 100% - 98.84% = 1.16%
```{r}
prediction <- predict(model, validationData)
confusionMatrix(validationData$classe, prediction)
```

# Evaluation  
Use the model to predict a vector of outcomes for the original test dataset
```{r}
finalPrediction <- predict(model, testing[, -length(names(testing))]) #remove the problem_id field from the prediction

```
The final Predictions were subsequently compared to the original test dataset.  They correctly predicted the outcome in all 20 cases.
